---
title: '**Clustering for Mall Customer data**'
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{float}
- \floatplacement{figure}{H}
- \usepackage{bm}
- \usepackage{longtable}
- \usepackage{booktabs}
- \usepackage{caption}
- \captionsetup[figure]{font=scriptsize}
- \captionsetup[table]{font=scriptsize}
vignette: |
  %\VignetteEngine{knitr::rmarkdown} %\VignetteIndexEntry{Foo}
urlcolor: blue
geometry: left=1cm,right=1cm,top=1cm,bottom=0.8cm
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(xtable.comment=FALSE)
```

 The Mall customers data set (https://www.kaggle.com/datasets/kandij/mall-customers) contains some basic data about your customers like Customer ID, age, gender, annual income and spending score. In this project, we try to cluster the customers based on the basic data. 
 
 

```{r echo=FALSE, message=FALSE}
customer_data =  read.csv("Mall_Customers.csv")
customer_data = na.omit(customer_data)
names(customer_data) = c("CustomerID", "Gender", "Age", "Annual_Income", "Spending_Score") 
customer_data=customer_data[,-1]
customer <- within(customer_data, {
  Gender_lab<- factor(Gender, labels  = 1:2, levels = c("Male","Female"))
})
customer=customer[,-1]
attach(customer)
```

- Summary Statistics

```{r echo=FALSE,results='asis'}
library(xtable)
xtable(summary(customer[,-4]),caption = "Summary Statistics",digits = 2)
```
 - Matrix scatterplot and Correlation matrix


```{r echo=FALSE,warning=FALSE,message=FALSE,fig.align="center",out.width = "70%",fig.cap="Matrix scatterplot"}
library(PerformanceAnalytics)
chart.Correlation(customer[,-4], histogram = TRUE,)
#str(customer)
``` 

```{r echo=FALSE,warning=FALSE,message=FALSE,fig.align="center",out.width = "70%",fig.cap="Box Plots"}

features <- colnames(customer)[-4]

par(mfrow = c(1, 3))  # Set the layout of the subplots

for (i in 1:length(features)) {
  feature <- features[i]
  boxplot(customer[,feature] ~ Gender_lab, data = customer, main = feature, xlab = 'Gender', ylab = feature)
}
```

```{r echo=FALSE,results='asis'}
df=cor(customer[,-4])
xtable(df, caption = "Correlation matrix")
``` 

- As the correlation between most of the variables are very low, therefore, rather than going with the correlation-based distance, we can use metric-based distance for clustering.
- Also I would suggest standardizing the variables as they are in different scales and some have very high ranges.
- There is not much variation of the variables when the Gender is condisered.

**Hierarchical Clustering with Complete linkage**

```{r, echo=FALSE,fig.align="center",out.width="60%",fig.cap="Hierarchical Clustering with Complete linkage"}
customer.sc <- scale(customer[,-4])
customer.complete <- hclust(dist(customer.sc), method = "complete")
plot(customer.complete, xlab = "", sub = "", 
	cex = 0.7)
abline(h=4
       ,col="red")##cut the tree
```
```{r, echo=FALSE, results='asis'}
clust<-cutree(customer.complete, 3)
t<-table(clust)
xtable(t(t), caption = "Number of observations within each cluster (Hierarchical Clustering)")
customer.new<-cbind(customer[,-4],clust)
c1<-apply(customer.new[clust==1,],2,mean)
c2<-apply(customer.new[clust==2,],2,mean)
c3<-apply(customer.new[clust==3,],2,mean)
d<-as.data.frame(rbind(c1,c2,c3))
row.names(d)<-c("Cluster 1","Cluster 2","Cluster 3")
xtable(d,caption = "Cluster means of the variables (Hierarchical Clustering)",table.placement="H",digits = 3)
```

```{r,echo=FALSE,fig.align="center",warning=FALSE,fig.cap="Matrix scatter plots"}
# Plot data
customer_sub=customer.new[,-4]
par(oma=c(4,1,1,1))
plot(customer.new[,-4],upper.panel=NULL,cex.main=0.8,col = c("red","blue","green")[customer.new$clust],pch=c(1,1,1)[customer.new$clust])

par(fig=c(0,1,0,1),oma=c(0,0,0,0),mar=c(0,0,0,0),new=TRUE)
plot(0,0,type="n",bty="n",xaxt="n",yaxt="n")


legend("top", legend=c("Cluster 1", "Cluster 2","Cluster 3"), 
       col = c("red","blue","green") ,pch=1,xpd = TRUE, horiz = TRUE,cex = 0.7)
par(mar=c(5.1,4.1,4.1,2.1))

```

```{r, echo=FALSE, results='asis'}
clust<-cutree(customer.complete, 4)
t<-table(clust)
xtable(t(t), caption = "Number of observations within each cluster (Hierarchical Clustering)")
customer.new<-cbind(customer[,-4],clust)
c1<-apply(customer.new[clust==1,],2,mean)
c2<-apply(customer.new[clust==2,],2,mean)
c3<-apply(customer.new[clust==3,],2,mean)
c4<-apply(customer.new[clust==4,],2,mean)
d<-as.data.frame(rbind(c1,c2,c3,c4))
row.names(d)<-c("Cluster 1","Cluster 2","Cluster 3","Cluster 4")
xtable(d,caption = "Cluster means of the variables (Hierarchical Clustering)",table.placement="H",digits = 3)
```

```{r,echo=FALSE,fig.align="center",warning=FALSE,fig.cap="Matrix scatter plots"}
# Plot data
customer_sub=customer.new[,-4]
par(oma=c(4,1,1,1))
plot(customer.new[,-4],upper.panel=NULL,cex.main=0.8,col = c("red","blue","green","yellow")[customer.new$clust],pch=c(1,1,1,1)[customer.new$clust])

par(fig=c(0,1,0,1),oma=c(0,0,0,0),mar=c(0,0,0,0),new=TRUE)
plot(0,0,type="n",bty="n",xaxt="n",yaxt="n")


legend("top", legend=c("Cluster 1", "Cluster 2","Cluster 3","Cluster 4"), 
       col = c("red","blue","green","yellow") ,pch=1,xpd = TRUE, horiz = TRUE,cex = 0.7)
par(mar=c(5.1,4.1,4.1,2.1))

```

**K-means clustering with k = 3**

 

```{r, echo=FALSE,results='asis'}
set.seed(100)
km.out <- kmeans(customer.sc, 3, nstart = 20)
t1<-table(km.out$cluster)
xtable(t(t1), caption = "Number of observations within each cluster (k=3)")
customer.new<-cbind(customer,clust)
k1<-apply(customer_sub[km.out$cluster==1,],2,mean)
k2<-apply(customer_sub[km.out$cluster==2,],2,mean)
k3<-apply(customer_sub[km.out$cluster==3,],2,mean)

d<-as.data.frame(rbind(k1,k2,k3))
row.names(d)<-c("Cluster 1","Cluster 2","Cluster 3")
xtable(d,caption = "Cluster means of the variables for k-means clustering (k=3)",table.placement="H",digits = 3)
kclust<-km.out$cluster
customer.new<-cbind(customer.new,kclust)
```
```{r,echo=FALSE,fig.align="center",warning=FALSE,fig.cap="Matrix scatter plots",out.width="80%"}
# Plot data
par(oma=c(4,1,1,1))
plot(customer_sub,upper.panel=NULL,cex.main=0.8,col = c("red","blue","green")[customer.new$kclust],pch=c(1,1,1)[customer.new$kclust])

par(fig=c(0,1,0,1),oma=c(0,0,0,0),mar=c(0,0,0,0),new=TRUE)
plot(0,0,type="n",bty="n",xaxt="n",yaxt="n")


legend("top", legend=c("Cluster 1","Cluster 2","Cluster 3"), 
       col = c("red","blue","green") ,pch=1,xpd = TRUE, horiz = TRUE,cex = 0.7)
par(mar=c(5.1,4.1,4.1,2.1))

```

- Age is a good variable to cluster the data into 3 clusters.

 **K-means clustering with k = 4**

```{r, echo=FALSE,results='asis'}
set.seed(100)
km.out <- kmeans(customer.sc, 4, nstart = 20)
t1<-table(km.out$cluster)
xtable(t(t1), caption = "Number of observations within each cluster (k=4)")
customer.new<-cbind(customer,km.out$cluster)
k1<-apply(customer_sub[km.out$cluster==1,],2,mean)
k2<-apply(customer_sub[km.out$cluster==2,],2,mean)
k3<-apply(customer_sub[km.out$cluster==3,],2,mean)
k4<-apply(customer_sub[km.out$cluster==4,],2,mean)
d<-as.data.frame(rbind(k1,k2,k3,k4))
row.names(d)<-c("Cluster 1","Cluster 2","Cluster 3","Cluster 4")
xtable(d,caption = "Cluster means of the variables for k-means clustering (k=4)",table.placement="H",digits = 3)
```

```{r,echo=FALSE,fig.align="center",warning=FALSE,fig.cap="Matrix scatter plots"}
kclust<-km.out$cluster
customer.new<-cbind(customer.new,kclust)
# Plot data
par(oma=c(4,1,1,1))

plot(customer_sub,upper.panel=NULL,cex.main=0.8,col = c("red","blue","green","yellow")[customer.new$kclust],pch=c(1,1,1,1)[customer.new$kclust])

# legend("top", col = mycols, legend = levels(d$Group1), pch = 20, 
#        xpd = NA, ncol = 3, bty = "n", inset = 0.01, pt.cex = 1.5)

par(fig=c(0,1,0,1),oma=c(0,0,0,0),mar=c(0,0,0,0),new=TRUE)
plot(0,0,type="n",bty="n",xaxt="n",yaxt="n")

legend("top", legend=c("Cluster 1","Cluster 2","Cluster 3","Cluster 4"), 
       col = c("red","blue","green","yellow") ,pch=1,xpd = TRUE, horiz = TRUE,cex = 0.7)
par(mar=c(5.1,4.1,4.1,2.1))
```

**K-means clustering with k = 5**

```{r, echo=FALSE,results='asis'}
set.seed(100)
km.out <- kmeans(customer.sc, 5, nstart = 20)
t1<-table(km.out$cluster)
xtable(t(t1), caption = "Number of observations within each cluster (k=4)")
customer.new<-cbind(customer,km.out$cluster)
k1<-apply(customer_sub[km.out$cluster==1,],2,mean)
k2<-apply(customer_sub[km.out$cluster==2,],2,mean)
k3<-apply(customer_sub[km.out$cluster==3,],2,mean)
k4<-apply(customer_sub[km.out$cluster==4,],2,mean)
k5<-apply(customer_sub[km.out$cluster==5,],2,mean)
d<-as.data.frame(rbind(k1,k2,k3,k4,k5))
row.names(d)<-c("Cluster 1","Cluster 2","Cluster 3","Cluster 4","Cluster 5")
xtable(d,caption = "Cluster means of the variables for k-means clustering (k=5)",table.placement="H",digits = 3)
```

```{r,echo=FALSE,fig.align="center",warning=FALSE,fig.cap="Matrix scatter plots"}
kclust<-km.out$cluster
customer.new<-cbind(customer.new,kclust)
# Plot data
par(oma=c(4,1,1,1))
plot(customer_sub,upper.panel=NULL,cex.main=0.8,col = c("red","blue","green","yellow","orange")[customer.new$kclust],pch=c(1,1,1,1,1)[customer.new$kclust])

par(fig=c(0,1,0,1),oma=c(0,0,0,0),mar=c(0,0,0,0),new=TRUE)
plot(0,0,type="n",bty="n",xaxt="n",yaxt="n")


legend("top", legend=c("Cluster 1","Cluster 2","Cluster 3","Cluster 4","Cluster 5"), 
       col = c("red","blue","green","yellow","orange") ,pch=1,xpd = TRUE, horiz = TRUE,cex = 0.7)
par(mar=c(5.1,4.1,4.1,2.1))
```
**K-means clustering with k = 6**

```{r, echo=FALSE,results='asis'}
set.seed(100)
km.out <- kmeans(customer.sc, 6, nstart = 20)
t1<-table(km.out$cluster)
xtable(t(t1), caption = "Number of observations within each cluster (k=4)")
customer.new<-cbind(customer,km.out$cluster)
k1<-apply(customer_sub[km.out$cluster==1,],2,mean)
k2<-apply(customer_sub[km.out$cluster==2,],2,mean)
k3<-apply(customer_sub[km.out$cluster==3,],2,mean)
k4<-apply(customer_sub[km.out$cluster==4,],2,mean)
k5<-apply(customer_sub[km.out$cluster==5,],2,mean)
k6<-apply(customer_sub[km.out$cluster==6,],2,mean)
d<-as.data.frame(rbind(k1,k2,k3,k4,k5,k6))
row.names(d)<-c("Cluster 1","Cluster 2","Cluster 3","Cluster 4","Cluster 5","Cluster 6")
xtable(d,caption = "Cluster means of the variables for k-means clustering (k=6)",table.placement="H",digits = 3)
```

```{r,echo=FALSE,fig.align="center",warning=FALSE,fig.cap="Matrix scatter plots"}
kclust<-km.out$cluster
customer.new<-cbind(customer.new,kclust)
# Plot data
par(oma=c(4,1,1,1))
plot(customer_sub,upper.panel=NULL,cex.main=0.8,col = c("red","blue","green","yellow","orange","purple")[customer.new$kclust],pch=c(1,1,1,1,1,1)[customer.new$kclust])

par(fig=c(0,1,0,1),oma=c(0,0,0,0),mar=c(0,0,0,0),new=TRUE)
plot(0,0,type="n",bty="n",xaxt="n",yaxt="n")


legend("top", legend=c("Cluster 1","Cluster 2","Cluster 3","Cluster 4","Cluster 5", "Cluster 5"), 
       col = c("red","blue","green","yellow","orange","purple") ,pch=1,xpd = TRUE, horiz = TRUE,cex = 0.7)
par(mar=c(5.1,4.1,4.1,2.1))
```



```{r, echo=FALSE, warning=FALSE,message=FALSE,fig.cap="Plot of number of clusters vs total within cluster sum of squares", fig.align='center', out.width="80%"}
library(factoextra)

#create plot of number of clusters vs total within sum of squares
fviz_nbclust(customer_sub, kmeans, method = "wss")
```